---
title: "A space-time-age model for district-level under-5 mortality estimation in India"
author: "Nathaniel Henry"
date: "`r Sys.Date()`"
documentclass: "`r conf$v$tex_style[[chapter_type]]`"
bibliography: "`r conf$v$bibliography`"
csl: "`r paste0(conf$v$citation_style, '.csl')`"
link-citations: true
---

```{r set_captions, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "", fig.align = 'center', echo=FALSE)

# Set caption text
captions <- list(
  fig1 = "Example caption for a figure.",
  t1 = "Example caption for a table."
)
```

# Abstract

\lipsum[1]


# Introduction

\lipsum[1-3]


## Household surveys as a tool for health system planning

\lipsum[4-6]


## Under-5 mortality across Indian states: progress and inequality

\lipsum[7-9]

## The need for local estimation

\lipsum[10-11]


# Methods

\lipsum[12-15]


## Geo-locating complete birth history data


\lipsum[1-2]


## Space-time-age mortality estimation model

\lipsum[3-9]


### Bias correction for retrospective survey data

\lipsum[10-11]


## Mortality forecasting

\lipsum[12-15]


# Results

\lipsum[1-2]


## District-level mortality and rankings

\lipsum[3-6]


## Comparison to 2025 and 2030 mortality targets

\lipsum[7-8]


# Discussion

\lipsum[9-10]


## Scale and inequality in mortality

\lipsum[11-14]


## Relationship with determinants of child health

\lipsum[1-2]


## Limitations of survey-based spatial mortality mapping

\lipsum[3-8]


### Uncertainty in rankings

\lipsum[9-10]


### Uncertainty in forecasting

\lipsum[11-12]


## Conclusions

\lipsum[1]

This is an example paragraph. To conclude, let's reference Figure \@ref(fig:fig1) and Table \@ref(tab:t1). Here are some example citations [@Aalbers; @Aguirre2009].


# References

<div id="refs"></div>
\newpage


# Figures and Tables

```{r fig1, out.width="70%", fig.cap = captions$fig1}
knitr::include_graphics(graphics_fps$fig1)
```
\newpage

```{r t1}
data_list$t1 %>%
  dplyr::mutate_all(linebreak, align='l') %>%
  knitr::kable(
    'latex', booktabs = T, escape = F, caption = captions$t1,
    col.names = linebreak(colnames(data_list$t1))
  ) %>%
  kable_styling(latex_options = "striped")
```
