---
title: "Introduction"
author: "Nathaniel Henry"
date: "`r Sys.Date()`"
documentclass: "`r conf$v$tex_style[[chapter_type]]`"
bibliography: "`r conf$v$bibliography`"
csl: "`r paste0(conf$v$citation_style, '.csl')`"
link-citations: true
---

```{r set_captions, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "", fig.align = 'center', echo=FALSE)

# Set caption text
captions <- list(
  fig1 = "Example caption for a figure.",
  t1 = "Example caption for a table."
)
```


# Abstract

\lipsum[1]


# Health system performance: the urgent need for better data

In 1948, the United Nations Universal Declaration on Human Rights asserted the fundamental and universal right of all people to "a standard of living adequate to the health and well-being of himself and his family, including [...] medical care and necessary social services[@UNGeneralAssembly1948]." However, more than 80 years after this declaration, a person's opportunities to live a long life in good health vary vastly depending on where they live. Gaps in health between countries and regions of the world are well-documented. However, health inequalities also manifest themselves within countries and even within walking distance. They appear in the earliest years of life: in Kano state in northern Nigeria, children are 2.5 times more likely to die before their fifth birthday than children born in the capital, Lagos[@Burstein2019]. They are also apparent in high-resource settings, such as urban counties in the United States, where life expectancy for men has been shown to vary by nearly 20 years across neighborhoods in a single U.S. county[@Dwyer-Lindgren2017]. The multifarious barriers to human flourishing that cause these inequalities at both the local and international level are both ethically troubling and a massive challenge to global health governance[@Ruger2006].

These challenges naturally raise questions about which institutions have the responsibility and capability to provide lifesaving services and remove barriers to human health. The discourse around health on the international stage has shifted over the past 20 years to a conception of "global health," where multilateral institutions and funders have greater perceived agency to coordinate health interventions[@Brown2006]. However, while international funders have successfully coordinated responses to acute mortality and disease threats over the past 20 years, these same experiences have also demonstrated that sustainable health services can only be delivered by health systems that are led by national stakeholders and operated locally[@WorldHealthOrganization2007; @WorldHealthOrganization2010]. Health systems can be defined in terms of their human resources and material components: they are driven by health care workers who rely on a financial and material infrastructure that is, in turn, managed by a constellation of planners, financial intermediaries and governing institutions[@Roberts2008]. They can also be defined in terms of their key operations: the World Health organization (WHO) lists service delivery, health workforce, information, medical products, vaccines and technologies, financing, and leadership and governance as the six core building blocks that constitute a functioning health system[@WorldHealthOrganization2007].

While health systems necessarily develop in the context of local conditions and priorities, they share the unifying aim of improving the health of the people they serve. Therefore, any attempt to manage or improve healthcare must be measured against its potential impact on health outcomes[@Roberts2008]. From local hospitals to national ministries of health, health policymakers need to make decisions about efficiently allocating funding, prioritizing at-risk groups, identifying and responding to health crises, and implementing long-term policy development and reform. All of these decisions require a consensual understanding among parties of conditions on the ground &mdash; that is, they require data examining the potential impact of decisions on health outcomes[@AbouZahr2015]. Without data on health outcomes, other sociological and economic analyses can only describe, not drive, health policy[@Roberts2008].

Complementary to their service provision activities, both international agencies and national bodies operate data collection and statistics systems that are designed to reveal actionable information about the state of health in a country. Many national ministries of health and statistics maintain Civil Registration and Vital Statistics (CRVS) operations to systematically register vital events such as birth and deaths, as well as surveillance systems for notifiable infectious diseases such as HIV, tuberculosis, and measles. In high-resource countries, largely complete and high-quality health information systems facilitate epidemiological investigation and decision-making. In many lower-resource settings, where the majority of childhood deaths and disease burden are concentrated, CRVS systems may be absent or incomplete, while infectious disease surveillance may be hindered by low completeness and reporting lags. To alleviate this data gap, international health institutions often fund household surveys that systematically collect information about key aspects of health. These surveys can be further supplemented by health modeling approaches: notably, modern spatial statistical modeling can reveal local inequalities that may fall below the sampling frame of the original survey[@Diggle2016].

This thesis asserts that high-quality CRVS and infectious disease surveillance are irreplaceable as a foundation for responsive health decision-making, and that these systems are therefore essential prerequisites for delivering sustainable and equitable health services to all people. The following sections of this chapter introduce the operation of these health data systems as well as the spatial modeling approaches that have been designed to supplement them. Past scholarship on data governance in global health has expressed the concern that modeling approaches have quelled the demand for high-quality national health surveillance without offering the same insights. In conclusion to this chapter, and more expansively throughout this thesis, I offer a partial solution to this problem of data governance: a statistical modeling framework that robustly incorporates deficient health surveillance records to measure health outcomes, and in so doing both estimates bias in the health surveillance system and provides an incentive for its improvement.


# National health data sources: history and uses


## Civil registration and vital statistics (CRVS) systems

- PAR: How a CRVS system operates
  - Birth records:
    - Often legally mandated
  - Death records:
    - Certification
  - Records flow[@Setel2007]
  - Dissemination and use in country statistical systems[@UnitedNationsStatisticsDivision2014]
  - Passive system; if birth registration not automatic, there may be systematic exclusion of children from vulnerable groups[@Fisker2019; @Hernandez2012]

- PAR: CRVS as a human right
  - Benefits conferred with a birth certificate
  - Benefits conferred to family with a death certificate[@Setel2007]
  - Example: Mexico's dual system for birth registration

- PAR: Estimated completeness of CRVS systems to date
  - Stems from differences in institutional histories[@Blake1955]
  - As of 2004, fewer than 1 in 100 residents in Southeast Asia and fewer than 1 in 10 Africans were covered by any birth or death registration system[@Mahapatra2007]; today, enormous gaps in service still remain.
  - Compare maps of death registration coverage in 2000 and 2017 (all ages and under-5s)[@Roth2018]
  - Causes of death: studies of cause-of-death registration quality have found that over 30% of all registered deaths in Egypt, Saudi Arabia, Bolivia, and Iraq could not be assigned with
  certainty to even a broad cause-of-death grouping, compared to less than 10% of registered deaths with the same coding issues in Finland, Australia, and Ireland [@Johnson2021].
  - LMICs: COVID-19 disrupted CRVS collection in many countries that did not consider it an "essential" service[@AbouZahr2021]
  
- PAR: Implications for the future of CRVS as a data source
  - Rate of change is too slow


## Infectious disease surveillance

- PAR: High-income settings: notifiable infectious diseases
  - Notification triggers investigation, contact tracing
  - Weekly surveillance of potentially epidemic infectious disease such as influenza, notifiable infectious diseases such as mumps, and drug resistance[@Thacker1989; @WorldHealthOrganization2017]
  - Risk assessment and early warning systems[@Vlieg2017]


- PAR: Lower-income settings: national infectious disease control programs
  - Measles, HIV, TB, Malaria
  - Supplemented by international funding[@Mauch2010]
  - Data collection process:
     - Missing data from private care providers[@Uplekar2016]
     - may have gaps after years of delay
     - Possibly hindered by lab access
  - Reported case notifications may be driven by program funding and access rather than underlying burden[@Rood2019]



## Household surveys

- PAR: What are household surveys?
   - Using DHS and MICS as examples[@Corsi2012; @Khan2019]
     - DHS funding: USAID
     - MICS funding: UNICEF
   - Often considered the "gold standard" in countries with deficient vital records or infectious disease surveillance
   - Benefits: systematic, can be detailed, designed to be representative at national level
   - However, limited and sometimes changing topics, smaller sample size, and long time gap

- PAR: Collection process:
   - Timeline and funding:
     - Dhs
   - Survey design
     - Standards
     - Country input, eg. India
     - Often designed to be representative at the national or admin1 level, by urban/rural
   - Data collection from a trained survey team
     - Number of households per cluster, number of clusters per country
     - Post-processing
   - What gets publicly released:
     - Report with key indicators
     - Deidentified microdata
     - Sometimes, includes jittered spatial identifiers

- PAR: Limitations
   - Limited changing topics
   - Smaller sample size
   - Recall biases
   - Can also be biased based on survey team, interview style, etc
   - Long time gap
     - (note retrospective questions)

## Special cases

- PAR: Special cases: China and India
  - History in India
    - Development of large household sample surveys, with sample sizes rivaling CRVS[@Dandona2016].
    - Beginning of CRVS system in select areas, but still only about 75% complete[Kumar2019].
  - History in China
    - History of census-based estimation[@Banister2004]
    - Combination of regular household censuses in combination with maternal and child
      mortality monitoring[@He2016], and CRVS/infectious disease surveillance that is rapidly
      increasing in coverage[@Zeng2020; @Vlieg2017]


## Standards for quality and usability

- PAR: Standards for health data quality, based on UN recommendations:
  - Compulsory and universal: Legally mandated, accessible, and using standard definitions nationwide
  - Timely: Data available in time to be used for decision-making
  - Accurate information about the individual and vital event
  - Complete: No individuals or sub-populations missing
  - Confidential: Individuals not identifiable from publicly-available data[@UnitedNationsStatisticsDivision2014]
  - Another perspective: SCORE criteria: survey populations and health risks; count births, deaths, and causes of death; optimize health service data; review progress and performance; enable data use for policy and action [@WorldHealthOrganization2021]


## Relationship between health data availability and health system capacity

- PAR: Tension between long-term need for high-quality health surveillance and short-term incentives
  - Long term: CRVS and infectious disease surveillance convey a set of unique health and 
    social benefits on the covered country
  - Short-term:
     - Required ramp-up needed for health surveillance to be used as a primary data source
     - With health spending per capita extremely low in some countries, how to
       justify spending money on capacity-building for a health surveillance system
       that may not be usable for decades, if ever?
     - (Related: can we find costing estimates for a functioning CRVS system?)

- PAR: The role of funders and international organizations
  - Funders have traditionally seen CRVS and infectious disease surveillance systems as de nulla fonts of health statistics, not as goods in their own right. Tradtionally, an office was assigned to improve CRVS capacity, but given no funding or authority to incentivize improvements [@AbouZahr2015]
  - Difficult challenge: tying together reporting structures across multiple institutions and levels of government, as demonstrated by Nigeria's efforts to institute a mandatory death registration system[@Makinde2020].
  - Awkward position for funders: control of data collection and governance


# Spatial modeling as a tool for responsive health interventions

- PAR: What is spatial modeling?
  - Modeling: Formal statistical assumptions about the "data generation process"
  - Bayesian data analysis using hierarchical models, which allows for intuitive understandings
    of randomness and uncertainty as a product of incomplete information[@McElreath2016].
  - Raw data enters a model, fits underlying parameters, which are then used to predict
    outcomes for a wider set of locations (discrete or continuous), often with uncertainty

- PAR: Why use spatial predictive modeling to estimate variation in health?
  - Key advantages: "fill in the gaps" left by existing data sources, identify focal areas, inequalities
  - Key disadvantages: communicating uncertainty, potential misspecification, data limitations, interpretability


## Disease mapping and small-area estimation

- PAR: Intro
  - Both approaches traditionally extend generalized linear mixed-effects regressions to add
    a spatial component
  - Discrete versus continuous, but both offer statistical formalizations of Tobler's First
    Law of Geography: "Everything is related to everything else, but near things are more related than distant things[@Tobler1970]."
  - Small samples are drawn from an underlying risk surface
  - Decomposition of variation into covariate-predicted, spatially-correlated, and IID[@Riebler2016]
  - Around for a long time, exploded in the 1990s with the greater availability of statistical
    software and the development of new spatial modeling frameworks that were computationally 
    tractable[@Besag1991].

- PAR: Background for small area estimation
  - Key: a large number of discrete areal units, where the number of sampled individuals might
    be small for some units, and potentially predictive covariates are available for all
    spatial units.
  - The classic Fay-Herriot model formulation, published in 1979 and widely used thereafter,
    estimated per-capita income across small areal units under the assumption that all
    observations were sampled independently and share the same variance function, with no
    spatial dependence[@Fay1979]
  - Later models added a term for local autocorrelation across space, allowing spatial
    modelers of health to draw predictive information from the local neighborhood structures
    across a study area[@Besag1991].

- PAR: Background for continuous disease mapping
  - Key: Predicting across a continuous surface where variance-covariance depends on space
  - Geostatistical modeling originally used to find mineral deposits[@Oliver2010]
  - How to invert?
    - Traditionally challenging
    - SPDE approach: approximates this continuous surface using a mesh, allowing for
      a far more efficient fitting approach[@Lindgren2011;@Miller2020]

- PAR: Extensions
  - Space-time modeling[@Mercer2015]
  - Machine learning methods for covariate generation[@bhatt2017]
  - dasymmetric mapping[@Tatem2017]
  - Species distribution mapping[@Hay2013]
  - Other extensions of regression
  - hotspot analysis
  - point pattern analysis[@Banerjee2014]


## Applications

- PAR: Who does spatial modeling?


### Low- and middle-income countries

- PAR: Mapping disease indicators and risk factors from household surveys
  - Generalizable methods for mapping in low-resource settings based on household surveys [@Diggle2016]
  - Wide array of possibilities for mapping[@Pigott2015]
  - Examples:
    - Malaria: MAP[@Weiss2019;@Nguyen2019]
    - Childhood infectious diseases and risk factors: IHME [@Osgood-Zimmerman2018]
    - Spatial variation in health system capacity: KEMRI [@Maina2019]
    - Child mortality [@Wakefield2020]
  - Software widely available and runnable on a personal PC

- PAR: Milieu
  - Draw predictive power from "covariates" that are well-established
    - remote sensing technologies used to map short-term weather patterns as well as long-term land use and climate change trends[@Ericksen2011]
    - VGI from (eg) HOTOSM[@Thomson2019]
    - Modeled estimates, notably WorldPop (actually a modeled estimate)[@Tatem2017]

- PAR: Short-term program implications for this kind of data:
  - Target and design new surveys
  - Track the effect of scaling up interventions[@Diggle2016]
  - Develop new programs [@Pigott2015]

### High-income countries

- PAR:
  - Greater data fidelity allows for a wider range of approaches
  - Use case 1 - Estimation of health outcomes for very local areal units, including United States census tracts[@Dwyer-Lindgren2014; @Zhang2014]
  - Use case 2 - Risk factor detection using observational data [@Liu2019]
  - Use case 3 - Age-period-cohort models for cancer incidence [@Papoila2014]; demographic modeling over space, age group, and causes of death [@Dwyer-Lindgren2016]
  - Monthly and weekly mortality reports[@Weinberger2020a]

- PAR: Completeness assumption is often grounded on quality checks; however, this does not
  mean fidelity is perfect
  - Examples of data quality checks from SCORE report[@WorldHealthOrganization2021]
  - Counterexample: COVID-19 mortality reporting issues
    - Time lag
    - Negative deaths
    - Cause misreporting [@Weinberger2020a]

## Limitations

- PAR LIMITATION - Usefulness
  - Long-term: household surveys in conjunction with statistical modeling leave major
    gaps in terms of knowledge
  - Compare to standards: compulsory and universal, timely, accurate, complete, and confidential

- PAR: LIMITATION - Uncertainty and its communication
  - Uncertainty from a Bayesian perspective often expressed in terms of Uncertainty
    Intervals (UI)
  - In mapping terms: The output of Bayesian spatial modeling is a large number of "candidate maps" that represent possible explanations of the underlying data, where better explanations are more likely to be included[@Patil2011]. Mapping outputs typically show summaries, such as the mean and uncertainty, of each pixel or areal unit in a candidate map.
  - BUT: This uncertainty is difficult to communicate, especially across space.
    - In this way, maps create meaning and certainty
  - This question of finality is supported by past research on cartography in policymaking,
    which has demonstrated how maps can convey a sense of finality to unsettled questions
    in space: one can settle any doubts with an assertion that "it's there on the map"
    [@Elwood2006].
  - Policy processes do not necessarily incorporate uncertainty

- PAR: LIMITATION - model misspecification
  - All models built on assumptions
  - Circularity and "models on top of models"
  - Example: mortality estimates for infectious diseases: IHME vs. WHO[@Tichenor2020] differences largely in areas with little death registration data. Differences in modeling approaches, prior specifications, and processing of the few data sources available (VA) have huge implications for global health policymaking and financing. Larger issue on the local scale.

- PAR: LIMITATION - who benefits? Relationship between modeler, data collector, and country
  - Statistical modeling still has a high barrier to entry
  - May create perverse incentives[@Tichenor2020]
  - Concerns of data imperialism[@Marchais2020]
  - Data doubles create representations of a country's health and development status that it has no control over, taking away levers for action in decision-making around global health financing and action [@Cinnamon2020a]

- PAR: LIMITATION - even with advanced modeling techniques, some analyses are intractable
   - Example: COVID-19 death counts


# New hierarchical modeling techniques for mapping health outcomes using incomplete health surveillance data

- PAR: Creating incentives for improving health surveillance data
  - Extending spatial modeling to start closing the gap in the health data divide
  - Developing models where even deficient vital records can offer insights into the
    health of a country that are impossible to derive from other sources, offering an 
    incentive for improvement
    - Certainty of these models increases as quality of surveillance increases
  - Goal: create a virtuous cycle where health surveillance data can be used for
    health system strengthening, and better funding/functioning of health systems at the
    local level lead to improved health surveillance

- PAR: Theoretical grounding for this approach
  - Fundamentally, health surveillance is an important building block of a functioning
    health system[@Roberts2008]
  - In line with the geographic movement towards "critical cartography", this project
    recognizes that maps are inherently value-laden representations of reality: the
    assumptions and use cases underlying maps of health have implications for the direction
    of health policy and reform[@Crampton2006].
  - The spatial aspect of this modelling is important:
    - Within a CRVS system, the principles of universality, timeliness, accuracy, completeness, and confidentiality (as well as barriers to these) often correspond to local processes that can be assessed using a spatial approach
    - variation across a country can tell us important things about the overall quality
      of the data
    - Barriers to high-quality health data overlap with health service capacity issues at
      the local level -- improving one should improve both
  - Necessary but not sufficient for adequate health data governance in LMICs, complementary to legal and governmental frameworks[@Tiffin2019]

- PAR: Modeling benefits of this approach
  - Crosswalks common in statistical modeling using a simple correction term, but more
    nuanced approaches to joint estimation are becoming more common thanks to improved
    tools for estimation
  - Bayesian hierarchical modeling approach, where the data-generating process includes the generation of CRVS/surveillance data[@Schmertmann2018]
  - Estimate the biases or incompleteness of health surveillance by comparing to a 
    "gold standard" data source
  - Fill in the "gaps" provided by survey data using estimates from high-performing locations

- PAR: Programmatic benefits of this approach
  - Use case for health surveillance
  - Explicitly track changes in completeness over space and time
    - As with health outcomes, CRVS improvement should be targeted and data-driven
  - Spatial approach allows us to identify different levels of policy and funding impact:
    - National level
    - State level
    - Local level
  - Temporal modeling allows for identification of change and possible exemplars/barriers
    to improvement
  - Can also be used to relax the completeness assumption in high-income countries

- PAR: Translating goals into principles of model construction and collaboration
  - Understanding the data generating process means country-specific partnerships
    - Specific data environment
    - Specific history of CRVS development with spatial implications
  - Generalizing these specific models into templates that can be applied across a wide
    variety of contexts:
    - For example, the framework used to jointly model rare health events and surveillance
      completeness can be applied both to neonatal mortality (Mexico) and tuberculosis
      incidence (Uganda)
  - Open sourcing results
    - Modeling approach that brings its own limitations and potential for harm to the foreground [@Cinnamon2020a]
    - Recognizing that a steep learning curve for GIS and spatial modeling tools still inhibits wide adoption [@Shannon2018]


## Thesis structure

- PAR: Explanation of structure
  - Series of country-specific collaborations with implications for broader country
    data environments
  - Working from lowest availability to highest availability of spatial CRVS data, 
    increasing modeling possibilities

- PAR: India


- PAR: Mexico


- PAR: Uganda


- PAR: Italy
  - Relaxing the assumption about fidelity of CRVS data, specifically cause-specific
    mortality

- PAR: Themes
  - The coverage of health surveillance data enables small-population estimation that can inform responsive and equitable health policy
  - Jointly estimating local variation in disease burden and input data quality
  - Spatial statistical modeling:
    - Short-term opportunities to supplement developing health surveillance and household surveys
    - No alternative to complete, high-quality national health surveillance systems


This is an example paragraph. To conclude, let's reference Figure \@ref(fig:fig1) and Table \@ref(tab:t1). Here are some example citations [@Aalbers; @Aguirre2009].


# References

<div id="refs"></div>
\newpage


# Figures and Tables

```{r fig1, out.width="70%", fig.cap = captions$fig1}
knitr::include_graphics(graphics_fps$fig1)
```
\newpage

```{r t1}
data_list$t1 %>%
  dplyr::mutate_all(linebreak, align='l') %>%
  knitr::kable(
    'latex', booktabs = T, escape = F, caption = captions$t1,
    col.names = linebreak(colnames(data_list$t1))
  ) %>%
  kable_styling(latex_options = "striped")
```
