---
title: "Introduction"
author: "Nathaniel Henry"
date: "`r Sys.Date()`"
documentclass: "`r conf$v$tex_style[[chapter_type]]`"
bibliography: "`r conf$v$bibliography`"
csl: "`r paste0(conf$v$citation_style, '.csl')`"
link-citations: true
---

```{r set_captions, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "", fig.align = 'center', echo=FALSE)

# Set caption text
captions <- list(
  fig1 = "Example caption for a figure.",
  t1 = "Example caption for a table."
)
```


# Abstract

\lipsum[1]


# Health system performance: the urgent need for better data

\lipsum[2-6]


# National health data sources: history and uses

\lipsum[7-10]


## Household surveys

\lipsum[11-15]


## Civil registration and vital statistics (CRVS) systems

\lipsum[1-5]


## Health surveillance systems

\lipsum[6-10]


## Other sources

\lipsum[11-15]


# Local estimation drives responsive health interventions

\lipsum[1-3]


## Disease mapping and small-area estimation

\lipsum[4-8]


## Applications

\lipsum[9-10]


### Low- and middle-income countries

\lipsum[11-13]


### High-income countries

\lipsum[13-15]


## Limitations

\lipsum[1-4]


# New hierarchical modeling techniques for mapping health outcomes using incomplete health surveillance data

\lipsum[5-12]


# Thesis structure

\lipsum[1-3]

This is an example paragraph. To conclude, let's reference Figure \@ref(fig:fig1) and Table \@ref(tab:t1). Here are some example citations [@Aalbers; @Aguirre2009].


# References

<div id="refs"></div>
\newpage


# Figures and Tables

```{r fig1, out.width="70%", fig.cap = captions$fig1}
knitr::include_graphics(graphics_fps$fig1)
```
\newpage

```{r t1}
data_list$t1 %>%
  dplyr::mutate_all(linebreak, align='l') %>%
  knitr::kable(
    'latex', booktabs = T, escape = F, caption = captions$t1,
    col.names = linebreak(colnames(data_list$t1))
  ) %>%
  kable_styling(latex_options = "striped")
```
