---
title: "Introduction"
author: "Nathaniel Henry"
date: "`r Sys.Date()`"
documentclass: "`r conf$v$tex_style[[chapter_type]]`"
bibliography: "`r conf$v$bibliography`"
csl: "`r paste0(conf$v$citation_style, '.csl')`"
link-citations: true
---

```{r set_captions, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "", fig.align = 'center', echo=FALSE)

# Set caption text
captions <- list(
  fig1 = "Example caption for a figure.",
  t1 = "Example caption for a table."
)
```


# Abstract

\lipsum[1]


# Health system performance: the urgent need for better data

BLOCK QUOTE

- PAR: Urgency of health goals in an era of inequality


- PAR: Increasingly, health system capacity is at the forefront


- PAR: Health systems cannot function without the feedback mechanisms created by health data


- PAR: Introduction to this section




# National health data sources: history and uses

- PAR: Introduction to section



## Civil registration and vital statistics (CRVS) systems

- PAR: What is a CRVS system?


- PAR: Brief history of CRVS systems, into the modern era


- PAR: How a CRVS system operates


- PAR: Uses of CRVS for health system management


- PAR: CRVS as a human right


- PAR: Estimated completeness of CRVS systems to date


- PAR: Implications for the future of CRVS as a data source



## Health surveillance systems

- PAR: What is a health surveillance system?


- PAR: High-income settings: notifiable infectious diseases


- PAR: Lower-income settings: national infectious disease control programs
  - HIV
  - TB
  - Malaria

- PAR: How health surveillance is used


## Household surveys

- PAR: How are household surveys used to measure health?


- PAR: History and funding of household surveys as a method for health data collection


- PAR: Topics covered - how are they chosen?


- PAR: Limitations



## Other sources

- PAR: Sentinel surveillance sites


- PAR: Censuses


- PAR: Graveyard records


## Relationship between health data availability and health system capacity

- PAR: Standards for health data quality, based on UN recommendations


- PAR: Vicious cycle of low spending, low quality, low utility for health data


- PAR: Tension between long-term need for high-quality health surveillance and short-term incentives


- PAR: Funding sources for household surveys



# Spatial modeling as a tool for responsive health interventions

- PAR: What is spatial and space-time modeling?


- PAR: Why use spatial modeling to estimate variation in health?


- PAR: Benefits and limitations of spatial modeling


## Disease mapping and small-area estimation

- PAR: Intro


- PAR: Background for small area estimation


- PAR: Background for disease mapping


- PAR: Methodological disagreements


- PAR: Similarities


### Extensions

- PAR: Extensions and alternatives


## Applications

- PAR: Who does spatial modeling?


### Low- and middle-income countries

- PAR: Mapping disease indicators and risk factors from household surveys


- PAR: Milieu - remote sensing technologies, increased data from (eg) HOTOSM and WorldPop


- PAR: Program implications for this kind of data



### High-income countries

- PAR: Use case 1 - epidemic detection

- PAR: Use case 2 - risk factor identification

- PAR: Use case 3 - advanced demographic modeling

- PAR: Greater data fidelity allows for a wider range of approaches


## Limitations

- PAR: LIMITATION - Uncertainty and its communication


- PAR: LIMITATION - model misspecification


- PAR: LIMITATION - even with advanced modeling techniques, some analyses are intractable


- PAR: LIMITATION - more fundamentally, do these create perverse incentives?



# New hierarchical modeling techniques for mapping health outcomes using incomplete health surveillance data

- PAR: Creating incentives for improving health surveillance data


- PAR: Theoretical grounding for this approach


- PAR: Modeling benefits of this approach


- PAR: Programmatic benefits of this approach


- PAR: Translating goals into principles of model construction and collaboration


- PAR: Open sourcing results
 



# Thesis structure

- PAR: Explanation of structure


- PAR: India


- PAR: Mexico


- PAR: Uganda


- PAR: Italy


- PAR: Themes



This is an example paragraph. To conclude, let's reference Figure \@ref(fig:fig1) and Table \@ref(tab:t1). Here are some example citations [@Aalbers; @Aguirre2009].


# References

<div id="refs"></div>
\newpage


# Figures and Tables

```{r fig1, out.width="70%", fig.cap = captions$fig1}
knitr::include_graphics(graphics_fps$fig1)
```
\newpage

```{r t1}
data_list$t1 %>%
  dplyr::mutate_all(linebreak, align='l') %>%
  knitr::kable(
    'latex', booktabs = T, escape = F, caption = captions$t1,
    col.names = linebreak(colnames(data_list$t1))
  ) %>%
  kable_styling(latex_options = "striped")
```
