---
title: "Introduction"
author: "Nathaniel Henry"
date: "`r Sys.Date()`"
documentclass: "`r conf$v$tex_style[[chapter_type]]`"
bibliography: "`r conf$v$bibliography`"
csl: "`r paste0(conf$v$citation_style, '.csl')`"
link-citations: true
---

```{r set_captions, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "", fig.align = 'center', echo=FALSE)

# Set caption text
captions <- list(
  fig1 = "Example caption for a figure.",
  t1 = "Example caption for a table."
)
```


# Abstract

\lipsum[1]


# Health system performance: the urgent need for better data

BLOCK QUOTE

- PAR: Urgency of health goals in an era of inequality
  - Everyone shares the same fundamental rights to health
  - However, one's chances of surviving childhood and experiencing a long, healthy life vary vastly depending on where one lives
  - Across countries
  - Within countries
  - This inequality constitutes the largest moral crisis of our generation

- PAR: What set of actors is responsible for improving access to lifesaving health services?
  - REF WHO and UN policy briefs
  - Governance happens across a broad constellation of actors and funders
  - Ultimately, sustained health services are delivered by in-country service providers
  - Quality affected by funding, human resources, government management

- PAR: National health systems cannot function well without the feedback mechanisms facilitated by health surveillance data
  - How to run well and improve? DATA.
  - Examples:
     - Allocating funding
     - Identifying at-risk groups
     - Responding to health crises
     - Policy development and reform [@Roberts2008]
  - What is the alternative to a data-driven, outcome-oriented approach? (A: not good)

- PAR: Introduction to this section
  - Introduction to two data types maintained and governed by national ministries of health and statistics: CRVS and infectious disease surveillance
     - Brief introduction to their development
     - Importance for national health policy and management
     - Current gaps between high-income and low-income countries
  - Introduction to other data types, primarily household surveys, that are intended to fill the gap left by deficient vital records
     - Deficiencies in relation to CRVS
  - Modeling, particularly spatial modeling, as a useful but imperfect tool for enhancing population-level estimates based on survey data
  - Concern that modeling is "crowding out" demand for health surveillance without offering the same services
  - Alternative: modeling work that uses deficient CRVS, estimating both completeness and providing a use case

# National health data sources: history and uses

- PAR: Introduction to section



## Civil registration and vital statistics (CRVS) systems

- PAR: What is a CRVS system?


- PAR: Brief history of CRVS systems, into the modern era
  - Death records in England
  - By what time did many countries have CRVS systems
  - Colonial CRVS systems
  - US CRVS system: 
  - Updated to electronic records
  - When did LMICs start their health surveillance systems?

- PAR: How a CRVS system operates
  - Birth records:
    - Often legally mandated
  - Death records:
    - Certification
  - Records flow (eg. United States)
  - Dissemination and use in country statistical systems

- PAR: Uses of CRVS for health system management
  - Described above
  - Examples:
    - Allocating funding
    - Identifying at-risk groups
    - Identifying health crises (opioids)
    - Health system reform

- PAR: CRVS as a human right
  - Benefits conferred with a birth certificate
  - Benefits conferred to family with a death certificate
  - Example: Mexico's dual system for birth registration

- PAR: Estimated completeness of CRVS systems to date
  - TODO: Compare maps of death registration coverage in 2000 and 2017 (all ages and under-5s)
  - Refer to WHO SCORE report
  - Causes?
  
- PAR: Implications for the future of CRVS as a data source
  - Rate of change is too slow


## Infectious disease surveillance

- PAR: What is a health surveillance system?


- PAR: High-income settings: notifiable infectious diseases
  - Notification triggers investigation, contact tracing
  - List of possible diseases


- PAR: Lower-income settings: national infectious disease control programs
  - Measles
  - HIV
  - TB
  - Malaria
  - Supplemented by international funding (see IHME Global Health Financing reports)
  - Data collection process:
     - incomplete
     - may have gaps after years of delay
     - Possibly hindered by lab access
     - Sometimes supplemented by new HMIS systems
  - Reported case notifications may be driven by program funding and access rather than underlying burden[@Rood2019]

- PAR: How health surveillance is used
    - Allocating funding
    - Identifying at-risk groups
    - Identifying health crises (opioids)
    - Health system reform


## Household surveys

- PAR: What are household surveys?
   - BRIEF history
   - Collection process
   - Often considered the "gold standard" in countries with deficient vital records or infectious disease surveillance
   - Benefits: systematic, can be detailed, designed to be representative at national level
   - However, limited and sometimes changing topics, smaller sample size, and long time gap

- PAR: Collection process
   - Using DHS and MICS as examples
   - Timeline and funding
   - Survey design
     - Standards
     - Country input, eg. India
     - Often designed to be representative at the national or admin1 level, by urban/rural
   - Data collection from a trained survey team
     - Number of households per cluster, number of clusters per country
     - Post-processing
   - What gets publicly released:
     - Report with key indicators
     - Deidentified microdata
     - Sometimes, includes jittered spatial identifiers

- PAR: Limitations
   - Limited and sometimes changing topics
   - Smaller sample size
   - Can also be biased based on survey team, interview style, etc
   - Long time gap
     - (note retrospective questions)


## Other sources

- PAR: Other sources
  - Graveyard records
    - Substitute for vital records, with a very limited coverage
    - How to correct for changing underlying population?
  - Sentinel surveillance sites
    - Longitudinal data on populations that may reflect wider national trends
    - Issue: may not be truly representative over time
  - Censuses:
    - Supposedly a full recounting of the population
    - Useful for demography
    - Even less information per person, larger time gaps

- PAR: Special cases: China and India
  - History in India
    - Development of large household sample surveys, with sample sizes rivaling CRVS
    - Beginning of CRVS system
  - History in China
    - History of census-based estimation
    - Combination of regular household censuses in combination with sentinal surveillance,
      child mortality monitoring, and CRVS/infectious disease surveillance that is rapidly
      increasing in coverage


## Standards for quality and usability

- PAR: Standards for health data quality, based on UN recommendations:
  - TODO



# Spatial modeling as a tool for responsive health interventions

- PAR: What is spatial modeling?
  - Modeling: Formal statistical assumptions about the "data generation process"
  - Raw data enters a model, fits underlying parameters, which are then used to predict
    outcomes for a wider set of locations (discrete or continuous), often with uncertainty
  - Geostatistical modeling originally used to find gold!
  - Fundamental tenet: Tobler's first law

- PAR: Why use spatial predictive modeling to estimate variation in health?
  - Key advantages: "fill in the gaps" left by existing data sources, identify focal areas, inequalities
  - Key disadvantages: communicating uncertainty, potential misspecification, data limitations, interpretability


## Disease mapping and small-area estimation

- PAR: Intro
  - Both approaches traditionally extend generalized linear mixed-effects regressions to add
    a spatial component
  - Decomposition of variation into covariate-predicted, spatially-correlated, and IID
  - Around for a long time, exploded in the 1990s with the greater convenience of hierarchical modeling techniques
  - Today, they take advantage of common statistical software and spatial data availability

- PAR: Background for small area estimation
  - Key: discrete areal units
  - County-level estimation in agriculture, then population
  - Health approaches back to the 1990s
  - Famous applications

- PAR: Background for continuous disease mapping
  - Key: predicting across a continuous surface where variance-covariance depends on space
  - How to invert?
    - Traditionally challenging
    - SPDE approach
  - Famous applications

- PAR: Similarities
  - The same data source might be amenable to either type of modeling
  - Continuous vs. discrete modeling, but both follow Tobler's law


### Extensions

- PAR: Extensions
  - Space-time modeling
  - Machine learning methods for covariate generation
  - Other extensions of regression

- PAR: Other spatial modeling applications in health
    - hotspot analysis
    - point pattern analysis
    - dasymmetric mapping


## Applications

- PAR: Who does spatial modeling?


### Low- and middle-income countries

- PAR: Mapping disease indicators and risk factors from household surveys
  - Examples: MAP, IHME, KEMRI
  - Software widely available and runnable on a personal PC

- PAR: Milieu
  - Draw predictive power from "covariates" that are well-established
    - remote sensing technologies
    - VGI from (eg) HOTOSM
    - WorldPop (actually a modeled estimate)

- PAR: Short-term program implications for this kind of data:
  - Used to produce estimates described at the front
  - Target and design new surveys
  - Concentrate new programs, funding, etc.


### High-income countries

- PAR: Use case 1 - epidemic detection
  - EuroMOMO

- PAR: Use case 2 - risk factor identification
  - Norwegian data (spatial study)

- PAR: Use case 3 - advanced demographic modeling
  - Projections of life expectancy

- PAR: Greater data fidelity allows for a wider range of approaches
  - Sub-population analyses
  - Forecasts
  - Monthly and weekly mortality reports

- PAR: Completeness assumption is often grounded on quality checks; however, this does not
  mean fidelity is perfect
  - Examples of data quality checks from SCORE report
  - Counterexample: COVID-19 mortality reporting issues
    - Time lag
    - Negative deaths
    - Cause misreporting


## Limitations

- PAR: LIMITATION - Uncertainty and its communication
  - Long-term: household surveys in conjunction with statistical modeling leave major
    gaps in terms of knowledge
  - Uncertainty from a Bayesian perspective often expressed in terms of Uncertainty
    Intervals (UI)
  - In mapping terms: the relative likelihood of different candidate maps
  - BUT: This uncertainty is difficult to communicate, especially across space
  - Policy processes do not necessarily incorporate uncertainty

- PAR: LIMITATION - model misspecification
  - All models built on assumptions
  - Circularity and "models on top of models"
  - Example: mortality estimates for infectious diseases: IHME vs. WHO

- PAR: LIMITATION - who benefits? Relationship between modeler, data collector, and country
  - Statistical modeling still has a high barrier to entry
  - May create perverse incentives
  - Concerns of data imperialism

- PAR: LIMITATION - even with advanced modeling techniques, some analyses are intractable
   - Example: COVID-19 death counts


# Relationship between health data availability and health system capacity

- PAR: Tension between long-term need for high-quality health surveillance and short-term incentives
  - Long term: CRVS and infectious disease surveillance convey a set of unique health and 
    social benefits on the covered country
  - Short-term:
     - Required ramp-up needed for health surveillance to be used as a primary data source
     - With health spending per capita extremely low in some countries, how to
       justify spending money on capacity-building for a health surveillance system
       that may not be usable for decades, if ever?
     - (Related: can we find costing estimates for a functioning CRVS system?)

- PAR: The role of funders and international organizations
  - Survey funders: awkward position with data governance
  - Programs run through a health system will encounter many levels of intermediaries
  - Tracking and accountabiliy becomes obscured
  - Same issue of short-term versus long-term payoff


# New hierarchical modeling techniques for mapping health outcomes using incomplete health surveillance data

- PAR: Creating incentives for improving health surveillance data
  - Extending spatial modeling to start closing the gap in the health data divide
  - Developing models where even deficient vital records can offer insights into the
    health of a country that are impossible to derive from other sources, offering an 
    incentive for improvement
    - Certainty of these models increases as quality of surveillance increases
  - Goal: create a virtuous cycle where health surveillance data can be used for
    health system strengthening, and better funding/functioning of health systems at the
    local level lead to improved health surveillance

- PAR: Theoretical grounding for this approach
  - Fundamentally, health surveillance is an important building block of a functioning
    health system
  - The spatial aspect of this modelling is important:
    - variation across a country can tell us important things about the overall quality
      of the data
    - Barriers to high-quality health data overlap with health service capacity issues at
      the local level -- improving one should improve both

- PAR: Modeling benefits of this approach
  - Crosswalks common in statistical modeling using a simple correction term, but more
    nuanced approaches to joint estimation are becoming more common thanks to improved
    tools for estimation
  - Data-generating process includes the generation of CRVS/surveillance data
  - Bayesian approach
  - Estimate the biases or incompleteness of health surveillance by comparing to a 
    "gold standard" data source
  - Fill in the "gaps" provided by survey data using estimates from high-performing locations

- PAR: Programmatic benefits of this approach
  - Use case for health surveillance
  - Explicitly track changes in completeness over space and time
    - As with health outcomes, CRVS improvement should be targeted and data-driven
  - Spatial approach allows us to identify different levels of policy and funding impact:
    - National level
    - State level
    - Local level
  - Temporal modeling allows for identification of change and possible exemplars/barriers
    to improvement
  - Can also be used to relax the completeness assumption in high-income countries

- PAR: Translating goals into principles of model construction and collaboration
  - Understanding the data generating process means country-specific partnerships
    - Specific data environment
    - Specific history of CRVS development with spatial implications
  - Generalizing these specific models into templates that can be applied across a wide
    variety of contexts:
    - For example, the framework used to jointly model rare health events and surveillance
      completeness can be applied both to neonatal mortality (Mexico) and tuberculosis
      incidence (Uganda)
  - Open sourcing results


# Thesis structure

- PAR: Explanation of structure
  - Series of country-specific collaborations with implications for broader country
    data environments
  - Working from lowest availability to highest availability of spatial CRVS data, 
    increasing modeling possibilities

- PAR: India


- PAR: Mexico


- PAR: Uganda


- PAR: Italy
  - Relaxing the assumption about fidelity of CRVS data, specifically cause-specific
    mortality

- PAR: Themes
  - The coverage of health surveillance data enables small-population estimation that can inform responsive and equitable health policy
  - Jointly estimating local variation in disease burden and input data quality
  - Spatial statistical modeling:
    - Short-term opportunities to supplement developing health surveillance and household surveys
    - No alternative to complete, high-quality national health surveillance systems


This is an example paragraph. To conclude, let's reference Figure \@ref(fig:fig1) and Table \@ref(tab:t1). Here are some example citations [@Aalbers; @Aguirre2009].


# References

<div id="refs"></div>
\newpage


# Figures and Tables

```{r fig1, out.width="70%", fig.cap = captions$fig1}
knitr::include_graphics(graphics_fps$fig1)
```
\newpage

```{r t1}
data_list$t1 %>%
  dplyr::mutate_all(linebreak, align='l') %>%
  knitr::kable(
    'latex', booktabs = T, escape = F, caption = captions$t1,
    col.names = linebreak(colnames(data_list$t1))
  ) %>%
  kable_styling(latex_options = "striped")
```
