---
title: "Mapping the relationship between tuberculosis burden and case notifications in Uganda"
author: "Nathaniel Henry"
date: "`r Sys.Date()`"
documentclass: "`r conf$v$tex_style[[chapter_type]]`"
bibliography: "`r conf$v$bibliography`"
csl: "`r paste0(conf$v$citation_style, '.csl')`"
link-citations: true
---

```{r set_captions, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "!ht", out.extra = "", fig.align = 'center', echo=FALSE)

# Set caption text
captions <- list(
  fig1 = "Example caption for a figure.",
  t1 = "Example caption for a table."
)
```

# Abstract

\lipsum[1]


# Introduction

\lipsum[1-3]


## Tuberculosis (TB) mapping in high-burden settings

\lipsum[4-5]


## Archetypes of TB data availability

\lipsum[6-10]


## Prior evidence on the relationship between TB burden and case notifications

\lipsum[11-14]


# Methods

\lipsum[1-2]


## District-level TB prevalence mapping

\lipsum[3-5]


## Extrapulmonary TB correction

\lipsum[6]


## Smoothed time trend estimation using case notifications

\lipsum[7-12]


# Results

\lipsum[13]


## Estimating the TB prevalence to notification ratio (PNR) by district

\lipsum[1-4]


## Comparison with predictive covariates of TB burden and health systems access

\lipsum[5-10]


## Time trends of tuberculosis prevalence, 2010-2019

\lipsum[11-13]


# Discussion

\lipsum[14-15]


## Model-based adjustment of TB case notification rates

\lipsum[1-6]


## “Nowcasting” local variation in TB prevalence and uncertainty

\lipsum[7-8]


## Extension to other data contexts

\lipsum[9-11]


## Extension to TB-HIV coinfection and policy implications

\lipsum[12-15]


## Conclusions

\lipsum[1]

This is an example paragraph. To conclude, let's reference Figure \@ref(fig:fig1) and Table \@ref(tab:t1). Here are some example citations [@Aalbers; @Aguirre2009].


# References

<div id="refs"></div>
\newpage


# Figures and Tables

```{r fig1, out.width="70%", fig.cap = captions$fig1}
knitr::include_graphics(graphics_fps$fig1)
```
\newpage

```{r t1}
data_list$t1 %>%
  dplyr::mutate_all(linebreak, align='l') %>%
  knitr::kable(
    'latex', booktabs = T, escape = F, caption = captions$t1,
    col.names = linebreak(colnames(data_list$t1))
  ) %>%
  kable_styling(latex_options = "striped")
```
